---
title: "Figure 6. Ro/e tissue distribution analysis of glial subtypes (STARTRAC)."
date: "2026/9/16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Clean workspace
rm(list=ls())
# Load setting and functions
source("../script/stat_plot_functions.R")
# Set output directory
output_dir <- "./"
```

(a) Ro/e (ratio of observed to expected cell number) heatmap of the glial subtypes Oligodendrocytes, Microglia and Astrocytes across tissue*group combinations (groups: GF_D0, GF_D30, MMT_D30), computed with Startrac::calTissueDist (chi-square method) on the `Tissue_group` interaction as the cluster variable. Symbols indicate enrichment levels: +++ Ro/e > 2; ++ 1 < Ro/e <= 2; + 0.5 <= Ro/e <= 1; +/- 0 < Ro/e < 0.5; - Ro/e = 0. (b) Bubble plot of Ro/e with cell counts and significance symbols. The Ro/e matrix is exported as `RoE_table.csv`.

```{r Roe_analysis, echo=TRUE}
library(Startrac)
library(pheatmap)
library(Seurat)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

# Read Seurat object (must contain cell type annotation and group info)
sce <- readRDS("Brain_new.rds")  # adjust to your input file path

phe <- sce@meta.data

# Tissue*group interaction as cluster variable
phe$Tissue_group <- paste(phe$Tissue, phe$group, sep = "_")

phe <- phe %>% filter(group %in% group_order)
phe$group <- factor(phe$group, levels = group_order)

# Keep selected cell types (glial subtypes)
phe <- phe %>% filter(subtype %in% subtype_keep)
phe$subtype <- factor(phe$subtype, levels = subtype_keep)

Roe <- calTissueDist(phe, byPatient = FALSE, colname.patient = "orig.ident", colname.cluster = "Tissue_group", colname.tissue = "subtype", method = "chisq", min.rowSum = 0)

# Convert table to a plain numeric matrix
Roe <- unclass(Roe)

# Export Ro/e matrix as CSV
Roe_csv <- as.data.frame(Roe) %>% rownames_to_column("Tissue_Group")
write.csv(Roe_csv, file.path(output_dir, "RoE_table.csv"), row.names = FALSE, quote = FALSE)
```

## a. Ro/e heatmap with enrichment symbols

```{r Roe_heatmap_symbol, echo=TRUE}
# Symbol matrix for enrichment level display
symbol_mat <- matrix("", nrow = nrow(Roe), ncol = ncol(Roe), dimnames = dimnames(Roe))
symbol_mat[Roe > 2]               <- "+++"   # strong enrichment
symbol_mat[Roe > 1 & Roe <= 2]    <- "++"    # moderate enrichment
symbol_mat[Roe >= 0.5 & Roe <= 1] <- "+"     # mild enrichment / normal
symbol_mat[Roe > 0 & Roe < 0.5]   <- "+/-"   # mild depletion
symbol_mat[Roe == 0]              <- "-"     # absent

# Color breaks and palette (grey-blue gradient)
breaks <- c(0, 0.5, 1, 2, max(Roe))
colors <- c("#F1F1F1", "#C6DBEF", "#9ECAE1", "#6BAED6", "#2171B5")

# Heatmap with enrichment symbols
hm_symbol <- pheatmap(Roe, scale = "none", display_numbers = symbol_mat, number_color = "black", fontsize_number = 10, color = colorRampPalette(colors)(100), breaks = seq(min(breaks), max(breaks), length.out = 101), cluster_rows = TRUE, cluster_cols = TRUE, border_color = "grey85", fontsize = 10, cellwidth = 30, cellheight = 25)
hm_symbol
```

## b. Bubble plot with cell counts

```{r Roe_bubble, echo=TRUE}
# Bubble plot data: long format + cell counts
roe_long <- as.data.frame(Roe) %>% rownames_to_column("Tissue_Group") %>% pivot_longer(cols = -Tissue_Group, names_to = "CellType", values_to = "RoE")
cell_counts <- phe %>% count(Tissue_group, subtype, name = "Count") %>% dplyr::rename(Tissue_Group = Tissue_group, CellType = subtype)

plot_data <- roe_long %>%
  left_join(cell_counts, by = c("Tissue_Group", "CellType")) %>%
  mutate(
    Significance = case_when(
      RoE > 2 & Count > 50 ~ "***",
      RoE > 1.5 & Count > 30 ~ "**",
      RoE > 1.2 ~ "*",
      TRUE ~ ""
    ),
    log_count = log10(Count + 1)
  )

p_bubble <- ggplot(plot_data, aes(x = Tissue_Group, y = CellType)) +
  geom_point(aes(size = log_count, fill = RoE), shape = 21, color = "black", stroke = 0.5) +
  geom_text(aes(label = Significance), color = "white", size = 3, fontface = "bold") +
  scale_fill_gradient2(low = "#738CC5", mid = "white", high = "#50BDBE", midpoint = 1, limits = c(0, max(plot_data$RoE)), name = "Ro/e") +
  scale_size_continuous(name = "Log10(Count+1)", range = c(3, 15)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid.major = element_line(color = "grey90"), legend.position = "right") +
  labs(title = "Ro/e Enrichment with Cell Abundance (Tissue*Group vs CellType)", x = "Tissue*Group", y = "Cell Type")

print(p_bubble)
```

## c. Ro/e heatmap with numeric values

```{r Roe_heatmap_numeric, echo=TRUE}
# Heatmap with numeric values
hm_num <- pheatmap(Roe, scale = "none", display_numbers = TRUE, number_color = "black", number_format = "%.2f", color = colorRampPalette(c("#738CC5", "white", "#50BDBE"))(100), cluster_rows = TRUE, cluster_cols = TRUE, fontsize = 10, fontsize_row = 10, fontsize_col = 10, border_color = "grey80", cellwidth = 30, cellheight = 25)
hm_num
```
